---
title: "EDA"
output: github_document
---

```{r, include = FALSE}
library(survivoR)
library(tidyverse)
library(stringr)
library(dplyr)
library(survival)
library(survminer)
library(gtsummary)
library(gridExtra)
library(RColorBrewer)
library(usmap)
```

```{r, warning = FALSE, message = FALSE}
## reading in data
survivor_data_final = 
  read.csv("data/survivor_data_final.csv")
```

### Summary

```{r, warning = FALSE, message = FALSE}
## ages and survival times
survivor_data_final %>% 
  select(age_during_show, 
         days_survived) %>% 
  tbl_summary(type = list(age_during_show ~ "continuous", 
                          days_survived ~ "continuous"), 
              statistic = list(all_continuous() ~ "{mean} ({sd})"), 
              label = list(c(age_during_show) ~ "Age During Show (Years)",
                           c(days_survived) ~ "Survival Time on Show (Days)")) %>% bold_labels()
``` 

Note: N = 728 refers to the total count of records (i.e. contestant occurrences) in `survivor_data_final`; distinct contestants may be listed in multiple records, across seasons and/or within seasons. 

```{r, warning = FALSE, message = FALSE}
## personality type
survivor_data_final %>% 
  group_by(personality_type_binary) %>% 
  summarize(n_personality_type = n_distinct(full_name),
            mean_days_survived = mean(days_survived)) %>% 
    na.omit()  %>%
  knitr::kable(digits = 3, col.names = c("Personality Type", "Distinct Persons", "Mean Days Survived"))

## POC
survivor_data_final %>% 
  group_by(poc) %>% 
  summarize(n_poc = n_distinct(full_name),
            mean_days_survived = mean(days_survived, na.rm = TRUE)) %>% 
    na.omit()  %>%
  knitr::kable(digits = 3, col.names = c("POC", "Distinct Persons", "Mean Days Survived"))


## gender
survivor_data_final %>% 
  group_by(gender) %>% 
  summarize(n_gender = n_distinct(full_name),
            mean_days_survived = mean(days_survived, na.rm = TRUE)) %>%
    na.omit()  %>%
  knitr::kable(digits = 3, col.names = c("Gender", "Distinct Persons", "Mean Days Survived"))
```

Note: We report distinct person counts by personality type, POC status, and gender.

### POC and Gender Representation Across Seasons

```{r, warning = FALSE, message = FALSE}
fill_color = brewer.pal(9,"Reds")[4]
survivor_poc_over_time = survivor_data_final %>% 
  group_by(version_season, poc) %>%
  summarize(count = n_distinct(full_name)) %>%
  mutate(freq = count / sum(count)) %>%
  filter(poc == "POC") %>%
  separate(col = version_season, into = c('NA', 'season'), sep = 2) %>%
  dplyr::select(-"NA") %>%
  mutate(season = as.numeric(season))

ggplot(data = survivor_poc_over_time, aes(x = season, y = freq, group = 1)) +
  geom_line(color = fill_color) +
  geom_point(color = fill_color) +
  geom_smooth(se = FALSE, color = "seashell") + 
  ggtitle("Change in Proportion of POC Over Time ") +
  xlab("Season") + ylab("Proportion of POC")
```

Note: We report distinct person counts by POC status.

```{r, warning = FALSE, message = FALSE}
fill_color = brewer.pal(9,"Purples")[5]
survivor_gender_over_time = survivor_data_final %>% 
  group_by(version_season, gender) %>%
  summarize(count = n_distinct(full_name)) %>%
  mutate(freq = count / sum(count)) %>%
  filter(gender == "Female") %>%
  separate(col = version_season, into = c('NA', 'season'), sep = 2) %>%
  dplyr::select(-"NA") %>%
  mutate(season = as.numeric(season))

ggplot(data = survivor_gender_over_time, aes(x = season, y = freq, group = 1)) +
  geom_line(color = fill_color) +
  geom_point(color = fill_color) +
  geom_smooth(se = FALSE, color = "purple") + 
  ggtitle("Change in Proportion of Women Over Time") +
  xlab("Season") + ylab("Proportion of Women")
```

Note: We report distinct person counts by gender.

### Concentration of Contestants by Age and Geography

```{r, warning = FALSE, message = FALSE}
fill_color = brewer.pal(9,"PuBuGn")[5]
ggplot(survivor_data_final, aes(x = age_during_show)) + 
  geom_histogram(aes(y = after_stat(count/sum(count))), bins = 15, fill = fill_color, col = "white") +
  scale_y_continuous(labels = scales::percent) + 
  ggtitle("Distribution of Ages Across Contestant Occurrences") +
  xlab("Age (During Show)") + ylab("Count")
```

Note: Since contestants can re-appear across seasons at different ages, we rely on discrete records from `survivor_data_final` (i.e. contestant occurrences) as the unit of analysis for this plot in order to ensure comprehensiveness of age data.

```{r, warning = FALSE, message = FALSE}
survivor_state = survivor_data_final %>%
  group_by(state) %>%
  summarize(n = n())

plot_usmap(
  data = survivor_state, values = "n", lines = "blue"
) + 
  scale_fill_continuous(type = "viridis", name = "Contestant Count", label = scales::comma
  ) + 
  labs(title = "US States", subtitle = "Geographic Distribution of Contestants") +
  theme(legend.position = "right")
```

Notes: 
(i) Seasons 2, 41, 42, and 43 have been removed from the exploratory analysis due to inconsistent number of days
(ii) Since contestants can re-appear across seasons with different states of residence, we similarly rely on discrete records from `survivor_data_final` (i.e. contestant occurrences) as the unit of analysis for this plot in order to ensure comprehensiveness of location data.