---
title: "Survival Analysis"
output: html_document
date: "2022-11-26"
---

```{r, include = FALSE}
library(survivoR)
library(tidyverse)
library(stringr)
library(dplyr)
library(survival)
library(survminer)
library(gtsummary)
library(riskRegression)
library(tab)
```

Statistical Analysis
Methodology
For our project, we interest using Survival Analysis. Suppose that there is a true survival time, T, as well as a true censoring time, C. The survival time represents the time at which the event of interest occurs: in this dataset, the time at which participant is voted out. The censoring time is the time at which the participant drop out of the game show or survived until the last day of the show.

We observed the Survival Time T and Censoring Time C. Suppose there is a random variable Y

$$Y = min(T,C)$$

In other words, if the event occurs before the censoring such that T $<$ C, then we observed the true survival time T. If censoring occurs before the event such as T $>$ C, then we observe the censoring time. The status indicator as, 

$$\delta = \begin{cases} 
      & 1 &  T\leq C \\
      & 0 & T > C
   \end{cases}$$

Thus, $\delta$ = 1 if we observe the true survival time, and $\delta$ = 0 if we observe the censoring.

## Survival Analysis


```{r}
## reading in data
survivor_data_final = 
  read.csv("data/survivor_data_final.csv")
```

## Create status and time variable
```{r}
# filtering of seasons moved to data wrangling file
# replaced all survivor_final with survivor_data_final since filtering was moved to use for EDA

status <- c()
time <- c()
for (i in 1:nrow(survivor_data_final))
  {
  if (survivor_data_final[i,17] == "Quit") {
    status[i] <- 0
    time[i] <- survivor_data_final$days_survived[i]
  }
  else if (survivor_data_final[i,17] == "Sole Survivor") {
    status[i] <- 0
    time[i] <- survivor_data_final$days_survived[i]
  } 
  else if (survivor_data_final[i,17] == "Runner-up") {
    status[i] <- 0
    time[i] <- survivor_data_final$days_survived[i]
  } else if (survivor_data_final[i,17] == "Co-runner-up") {
    status[i] <- 0
    time[i] <- survivor_data_final$days_survived[i]
  }
  else {
    status[i] <- 1
    time[i] <- survivor_data_final$days_survived[i]
  }
}
survivor_data_final[,25] <- status
survivor_data_final[,26] <- time

colnames(survivor_data_final)[25] <- "status"
colnames(survivor_data_final)[26] <- "time"
```

## Survival Unadjusted Model
```{r}
surv_model_unadj <- survfit(Surv(time, status) ~ 1)

plot(surv_model_unadj, xlab = "Days",
     ylab = "Estimated Probability of Survival")
```

## Cox-proposional hazard model slection using `selectCox`
```{r}
surv_model_full <- coxph(data = survivor_data_final,
                         Surv(time, status)~ poc + age_during_show + 
                           personality_type_binary)

selectCox(Surv(time, status)~ poc + age_during_show + personality_type_binary + Midwest, data = survivor_data_final)
```

## Cox-proposional hazard model 1
```{r}
surv_model_cox1 <- coxph(data = survivor_data_final,
  Surv(time, status) ~  poc + age_during_show + personality_type_binary)

summary(surv_model_cox1)

#Clean table for Cox PH
tabcoxph(surv_model_cox1, 
         var.labels = list(poc = "POC", age_during_show = "Age", personality_type_binary = "Personality Type", gender = "Gender"), 
         decimals = 2)
```
**HR interpretations:**

As age increases in years, there is no difference in survival since the hazard ratio is 1. We are 95% confident that the HR falls between 0.9922 to 1.008. However, the HR is not statistically significant since the p-value of 0.998 > 0.05. 

For the gender variable, the HR value of 0.96 would suggest that the survival time for Male contestants compared to Female contestants (the reference category) is longer. We are 95% confident that the HR falls between 0.82 and 1.12. However, the HR is not statistically significant since the p-value of 0.57 > 0.05. 

For the POC variable, the HR value of 0.8881 would suggest that the survival time for White contestants compared to POC contestants (the reference category) is longer. We are 95% confident that the HR falls between 0.7456 and 1.058. However, the HR is not statistically significant since the p-value of 0.184 > 0.05.

For personality type, the HR value of 1.0631 would suggest that the survival time for Introvert contestants compared to Extrovert contestants (the reference category) is shorter. We are 95% confident that the HR falls between 0.9076 - 1.245. However, the HR is not statistically significant since the p-value of 0.448 > 0.05. 

## Cox-proposional hazard model 2
```{r}
surv_model_cox2 <- coxph(data = survivor_data_final,
  Surv(time, status) ~  ethnicity + age_during_show + personality_type_binary)

summary(surv_model_cox2)
```

## Cox-proposional hazard model 3 - interaction
```{r}
surv_model_cox3 <- coxph(data = survivor_data_final,
  Surv(time, status) ~  poc * age_during_show * personality_type_binary)

summary(surv_model_cox3)
```


The survival curve, or survival function, is defined as
$$S(t) = Pr(T>t)$$

## Kaplan-Meier plotter-personality
```{r}
surv_model_per <- survfit(Surv(time, status)~ survivor_data_final$personality_type_binary)

ggsurvplot(
  surv_model_per,
  data = survivor_data_final,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Extrovert", "Introvert"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)

#basic graph
# plot(surv_model_per, xlab = "Days",
#      ylab = "Estimated Probability of Survival", col = c(2,4))
```

## Kaplan-Meier plotter-White vs Non-White
```{r}
surv_model_poc <- survfit(Surv(time, status)~ survivor_data_final$poc)

ggsurvplot(
  surv_model_poc,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("POC", "White"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```

## Kaplan-Meier plotter-gender
```{r}
surv_model_sex <- survfit(Surv(time, status)~ survivor_data_final$gender)

ggsurvplot(
  surv_model_sex,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("female", "male"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```
## Kaplan-Meier plotter-Region
```{r}
surv_model_Region <- survfit(Surv(time, status)~ survivor_data_final$region)

ggsurvplot(
  surv_model_Region,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Midwest", "Northeast", "South", "West"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```


## Kaplan-Meier plotter-Northeast
```{r}
surv_model_NE <- survfit(Surv(time, status)~ survivor_data_final$NE)

ggsurvplot(
  surv_model_NE,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Non-Northeast", "Northeast"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```

## Kaplan-Meier plotter-South
```{r}
surv_model_South <- survfit(Surv(time, status)~ survivor_data_final$South)

ggsurvplot(
  surv_model_South,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Non-South", "South"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```

## Kaplan-Meier plotter-West
```{r}
surv_model_West <- survfit(Surv(time, status)~ survivor_data_final$West)

ggsurvplot(
  surv_model_West,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Non-West", "West"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)

```
## Kaplan-Meier plotter-Midwest
```{r}
surv_model_Midwest <- survfit(Surv(time, status)~ survivor_data_final$Midwest)

ggsurvplot(
  surv_model_Midwest,
  data = survivor_data_final,
  size = 1,                 # change line size
  conf.int = FALSE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Non-Midwest", "Midwest"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```

## log-rank for gender

```{r}
fit.cox <- survfit(Surv(time, status)~ survivor_data_final$gender)
logrank.test <- survdiff(Surv(time, status) ~ survivor_data_final$gender)
logrank.test
#p = 0.5 >0.05 not statistically significant, no strong evidence to say there is difference in survival time between the female and male.

summary(fit.cox)$logtest[1]
summary(fit.cox)$waldtest[1]
summary(fit.cox)$sctest[1]

logrank.test$chisq
```

## log-rank for White vs Non-White
```{r}
fit.cox <- survfit(Surv(time, status)~ survivor_data_final$poc)
logrank.test <- survdiff(Surv(time, status) ~ survivor_data_final$poc)
logrank.test
```

## log-rank for Northeast
```{r}
fit.cox <- survfit(Surv(time, status)~ survivor_data_final$NE)
logrank.test <- survdiff(Surv(time, status) ~ survivor_data_final$NE)
logrank.test
```
